---
title: "lab_02"
author: "Sam"
date: "2024-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

For this lab we'll be working with a slice of Maryland voter registration data: records of people who have registration dates this year through Aug. 7. Let's load it.

```{r}
md_voters_2024 <- read_rds("data/lab_02.rds")
```

## Examine Data

Take a look at the data, getting familiar with what it contains. Most of the columns should be self-explanatory, but there are two date columns: STATE_REGISTRATION_DATE and COUNTY_REGISTRATION_DATE. The former represents the date the person registered as a Maryland voter; the latter is when they registered in their current county. They may be the same, but if a voter moves from one county to another they will differ. This is important because if the two dates are the same, that's a pretty good indication that this is a *new* voter, someone who has no history of registering in Maryland. You'll need to keep that definition in mind.

## Answer questions

**Q1** How many voters appear to be *new* registrants since July 21, when President Joe Biden announced he would leave the race, based on the information above? Create a dataframe to hold the only those voters. Which counties have seen the most? Take a look at county population figures here and compare the two in your answer: https://worldpopulationreview.com/us-counties/maryland.

##A1: There are 17083 new voters post July 21. The counties that have seen the most new registration are Montgomery and Baltimore county. The two that have seen the most increase are also two of the most populated counties in the state. Therefore, it makes sense that there is a large increase.

```{r}
md_voter_post_July21 <- md_voters_2024 |> 
filter(STATE_REGISTRATION_DATE > "2024-07-21") |> 
arrange(STATE_REGISTRATION_DATE)
```


```{r}
new_maryland_voter_by_count <- md_voter_post_July21 |> 
group_by(COUNTY) |>
    summarize(voter_count=n()) |> 
arrange(desc(voter_count))
```  

**Q2** A better question is whether the new voters you found in Q1 are an outlier compared to other weeks this year. Using the original dataframe with 2024 registrants, use `mutate()` and lubridate to produce week numbers from the COUNTY_REGISTRATION_DATE (Lubridate has a function that turns a date into a week number. Use that to answer this question). Save that to the original dataframe. Then calculate how many voters are in each week. Describe the distribution of voters groups in a sentence - are the weeks from July 21-August 7 different from the rest, and if so, how?

**A2**: A majority of new voter registration came in the weeks and months after Biden droped out, but the highest actually came in the week of June 10. This is probably because the summer seems like the most likely time for people to register. But two of the top most registered weeks were the direct weeks after Biden dropped out of the race, which basically describes what we already knew. Democratic engagement went up massively after Kamala Harris became the presumptive favorite.

```{r}
md_voters_week <- md_voters_2024 |> 
  mutate(
    COUNTY_REGISTRATION_DATE = as.Date(COUNTY_REGISTRATION_DATE, format = "2024-07-21"),
    week_number = week(COUNTY_REGISTRATION_DATE)
  )
```

```{r}
voters_per_week <- md_voters_week |> 
  group_by(week_number) |> 
  summarize(voters_count = n()) |> 
  arrange(week_number)
```

**Q3** Using lubridate, show which month saw the largest number of registrations. Does the answer help shed more light on Q2?
**A3** : Yes, this showed that July was easily the most popular. It has the highest margin between two of the closest months, as it had more than 6,000 voters in July than in June. I think this showed that a good amount of people were largely enthused by Kamala Harris, particularly young people, who are the highest demographic that is likely to register to vote.

```{r}
md_voters_month <- md_voters_2024 |> 
  mutate(
    COUNTY_REGISTRATION_DATE = as.Date(COUNTY_REGISTRATION_DATE, format = "2024-07-21"),
    month_number = month(COUNTY_REGISTRATION_DATE)
  )
```

```{r}
voters_per_month <- md_voters_month |> 
  group_by(month_number) |> 
  summarize(voters_count = n()) |> 
  arrange(month_number)
```

**Q4** Maryland has, officially, five regions: https://www.visitmaryland.org/article/maryland-regions. Modify the original dataframe again to add a `region` column based on the `COUNTY` value using case_when. Are you going to have to do this for every Maryland county plus Baltimore City? Yes, you are. Which regions and parties account for the most new voters?

**A4** : Central Maryland accounts for the most new voters. The Democrats in the Capital Region had more voter registration than any other political party in a region, which is interesting because it doesn't align. The Republicans saw the most registration in the Central Region. 

```{r}
md_voters_2024 <- md_voters_2024 |> 
  mutate(
    region = case_when(
      COUNTY %in% c("Allegany", "Garrett", "Washington") ~ "Western Maryland",
      COUNTY %in% c("Frederick", "Montgomery", "Prince George's") ~ "Capital Region",
      COUNTY %in% c("Anne Arundel", "Baltimore County", "Baltimore City", "Carroll", "Harford", "Howard") ~ "Central Maryland",
      COUNTY %in% c("Calvert", "Charles", "Saint Mary's") ~ "Southern Maryland",
      COUNTY %in% c("Caroline", "Cecil", "Dorchester", "Kent", "Queen Anne's", "Somerset", 
                    "Talbot", "Wicomico", "Worcester") ~ "Eastern Shore Maryland"))
```

```{r}
maryland_voters_region <- md_voters_2024 |> 
group_by(region) |>
    summarize(voter_count=n()) |> 
arrange(desc(voter_count))
```  

```{r}
voters_region_party <- md_voters_2024 |> 
group_by(region, PARTY) |>
    summarize(voter_count=n()) |> 
arrange(desc(voter_count))
```  
**Q5** What's an additional question you could ask of this data?

**A5** What percentage of voters to the population of the region registered? Because obviously, there are some counties that have larger populations, so it would be interesting to see how the smaller counties stack up with the larger counties.
