---
title: "Class Reference"
author: "Sam Jane"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).

```{r}
setwd("~/jour472/data_journalism_2024_fall")
```

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
To install packages, we use the function install.packages(). 
You only need to install a library once, the first time you set up a new computer to do data journalism work. You never need to install it again, unless you want to update to a newer version of the package.

```

### Group by count
```{r}
data |> function
data |> group_by(COLUMN NAME) |> summarize(VARIABLE NAME = AGGREGATE FUNCTION(COLUMN NAME))
You can also use the summarize function, which can create a new data frame

```

### Do the sum
```{r}
Adding up the sum of a group of numbers is relatively easy in 
Put the column we want to total – “amount” – inside the sum() function sum(amount)
EXAMPLE
maryland_winred_contributions |>
  group_by(memo_text) |>
  summarise(
    count_contribs = n(),
    total_amount = sum(amount)
  ) |>
  arrange(desc(total_amount))
```

HOW TO DO MEAN
-  count_contribs = n(),
    total_amount = sum(amount),
    mean_amount = mean(amount),
    median_amount = median(amount)

#MUTATE
- You can add a column by using the mutate function
EXAMPLE 
``` {r}
primary_24 |>
  select(office_name, office_district, candidate_name, party, county_name, election_day, votes) |>
  mutate(
  percent_election_day = election_day/votes
)
```

#HOW TO MAKE DATA MORE CLEAN
**Answer**: str_to_title(): This function converts the first letter of each word in a string to uppercase while converting the rest to lowercase, effectively title-casing the string. str_to_lower(): This function converts all characters in a string to lowercase.

#ASCEND ORDER
``` {r}
primary_24 |>
  select(office_name, office_district, candidate_name, party, county_name, election_day, votes) |>
  mutate(
  percent_election_day = (election_day/votes)*100
)  |> 
  arrange(percent_election_day)
```