title: "Reverse Engineering Project"
author: "Sam Jane, Loretta Pulwer, Katharine Wilson"
date: "October 3, 2024"
output:
  html_document: 
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
      
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```
## Introduction 

In this notebook, we are reverse engineering the story, [Why many Americans don't vote](https://projects.fivethirtyeight.com/non-voters-poll-2020-election/)

## Load libraries
```{r}
# Load the non voters data
non_voters <- read_csv("nonvoters_data.csv")
```
## Load and Cleaning Data

The source of the data was a Pew Research Center survey that asked Americans on a scale of one through four how likely they were to vote in elections. This also asked other follow up questions, such as how long do you wait in line, what's the probability you leave work, etc. Some of the problems with this data was the weighting of people, there was a certain weight column that influenced if we could count users listed as whole individual people.

## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Sentence text**: Of the 8,000-plus people we polled, we were able to match nearly 6,000 to their voting history. We analyzed the views of the respondents in that slightly smaller group, and found that they fell into three broad groups: 1) people who almost always vote; 2) people who sometimes vote; and 3) people who rarely or never vote. People who sometimes vote were a plurality of the group (44 percent), while 31 percent nearly always cast a ballot and just 25 percent almost never vote.
* **Analysis summary**: We found that most Americans are sporadic in their voting habbits, followed by Americans who always vote. Sporadic had the highest difference between the two, meaning that most Americans are going to vote in some elections, and not vote in others.

To get to this answer, we first grouped by the three categories of voters (always vote, sometimes vote, and rarely or never vote) and counted the voters in each category. We then found the percent of each category of the total respondents. Finally, we weighed the responses. 
```{r}
voter_responseQ26 <- non_voters |> 
group_by(voter_category) |>
summarize (count_voters = n())
```
```{r}
voter_responseQ26 <- non_voters |> 
  group_by(voter_category) |> 
  summarize(count_voters = n()) |> 
  mutate(percentage_voters = (count_voters / 5836) * 100)
```

```{r}
total_weight <- non_voters |>
filter(Q26 != -1) |>
summarize(total = sum(weight))
```

### Sentence 2

* **Sentence text**: Of the three groups of voters we identified, those voters who only vote some of the time were actually the likeliest to report having stood in line for more than an hour; they were also likelier than those who vote more regularly to say they couldn’t get off work to vote.”
* **Analysis summary**: We were able to confirm the finding, showing that nearly 21& of sporadic voters found themselves in line at a voter box for over an hour and nearly 13% said they could not vote because it was difficult to take work off. This confirmed the results of the study.

We first filtered for only people who answered yes to question 18 (have you or any members of your household ever waited in line for more than an hour to vote). We then grouped by the voter categories for these yes respondents. Finally, we found the percentages for voter categories (how often people vote) who said yes to this question. 

```{r}
# Load the non voters data
non_voters <- read_csv("nonvoters_data.csv")
```

```{r}
# Waiting in line ?
non_voters |>
  filter(Q18_8 == 1) |>
  group_by(voter_category) |>
  summarise(count = n())
```

```{r}
# Percentage
333/1811 *100
241/1451 *100
531/2574 *100
```

```{r}
# Who couldn't get off work?
non_voters |>
  filter(Q18_7 == 1) |>
  group_by(voter_category) |>
  summarise(count = n())
```

```{r}
# Percentage
121/1811 *100
186/1451 *100
251/2574 *100
```

### Sentence 3

* **Sentence text**: There are, of course, other systemic reasons why some people might vote more inconsistently. Our survey found, for instance, that occasional voters were slightly more likely than frequent voters to have a long-term disability (8 percent vs. 5 percent), and nonvoters were even more likely to fall into this category (12 percent).
* **Analysis summary**: Our numbers matched the Pew Research article, except they rounded up in their findings. This is interesting, although it makes sense, because having a disability would make it difficult to vote. There are easier options to vote that are starting to become more common, such as absentee ballots and early voting that could make voting in elections more accessible for people with disabilities.


We first filtered to only people who said they have a disability. We then grouped by their voter category to see how often people with disabilities vote. We then found the percentages for each category of voting habit. 
```{r}
# Disabilities 
non_voters |>
  filter(Q10_1 == 1) |>
  group_by(voter_category) |>
  summarise(count = n())
```

```{r}
# Percentage
88/1811 *100
168/1451 *100
213/2574 *100
```

### Sentence 4

* **Sentence text**: About 24 percent of Black respondents said that they had to stand in line for more than an hour while voting
* **Analysis summary**: We also found that around 1/4th of Black respondents said they had to stand in line for over an hour while voting. This is interesting because it shows how disparities in where polling places are located can harm specific races. 

To get to this answer, we filtered the data to make a data set with just Black voters. We then grouped by their responses and calculated the percent of how many said yes to the question. 

```{r}
# Filtering for just Black respondents 
Black_non_voters <- non_voters|>
  filter(race == "Black")
```

```{r}
# Group by if they stood in line for 1 hr +  
Black_non_voters |> 
  group_by(Q18_8)|>
  summarise(count=n())
```
```{r}
# Percentage of Black respondents standing in line for 1 hr + 
(248/932)*100
```

```{r}
non_voters|>
  group_by(race)|>
  summarise(count=n())
```
### Sentence 5
* **Sentence text**: “In the survey, we asked voters who have missed at least one national election — which included some people who almost always vote — why they didn’t cast a ballot...But another 31 percent said that they decided not to vote because they disliked the candidates.
* **Analysis summary**: We got the same percentage of people who said they decided not to vote because they disliked the candidates. This is interesting because it reflects on the impacts of our two party system. 

We found this answer by filtering to look only at the people who regularly or sporadicly vote. We then filtered by people who responded to Q29, which means that they have missed at least one national election. We then grouped by their answer to Q29_1 (which meant that they decided not to vote because they didn't like the candidate). We then got the percentage of that number from the people who have missed at least one national election who vote at least occasionally. 

```{r}
real_non_voters <- non_voters |> 
  filter(voter_category == "always"|voter_category == "sporadic")
```


```{r}
real_non_voters_filtered <- real_non_voters |> 
  filter(Q29_1 == 1 | Q29_2 == 1 | Q29_3 == 1 | Q29_4 == 1 | Q29_5 == 1 | Q29_6 == 1 | Q29_7 == 1 | Q29_8 == 1 | Q29_9 == 1 | Q29_10 == 1)
```


```{r}
real_non_voters_filtered |>
  group_by(Q29_1)|>
  summarise(count=n())
```
```{r}
(110 / 345)*100
```
