---
title: "lab_06"
author: "Sam Jane"
date: "2024-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Tabula

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse and any other libraries you'll need
options(scipen=999)
library(tidyverse)

```

## Get Our PDF

We'll be working with the [Maryland monthly voter registration update](https://elections.maryland.gov/pdf/vrar/2024_09.pdf) from the State Board of Elections. You'll want to download it to a place you'll remember (like your Downloads folder, or the labs folder in your repository). The goal is to write a couple of paragraphs that summarize the changes in voter registration in Maryland during the month of September.

## Setup Tabula

Start Tabula, then go to <http://127.0.0.1:8080/> in your browser. Click the "Browse" button and find the PDF file and click "open", and then click the "Import button" in Tabula. This will take a few seconds.

This PDF has multiple possible tables to extract. We're going to make four dataframes: 

* new registrations by party
* removals by party
* changes in registration
* current registrations by county.

You will need to draw boxes around four sections and export four CSV files that you will load into R. In Tabula, draw a box around the borders of those tables and click the "Preview & Export Extracted Data" button for each one. The export should look pretty clean, but might be missing some column headers. You'll need to include total columns.

Export each CSV (each one should be called `tabula-2024-09.csv` by default, so you should rename them for clarity) to your lab_06/data folder.

From there, you will need to read in and clean up the data, and in particular the headers. You can choose to include the headers from the PDF in your exported CSV files OR to exclude them and add them when importing. The pre_lab has an example of how to do this!

## Load and clean up the data in R

You will need to read in and clean up the data so that it can be used for analysis. By "clean" I mean the column headers must not contain spaces and they should have meaningful names, not "x1" or something similar. The current registrations dataframe *MUST* include the county name, but should not include the columns under `Changes` and `Party Affiliation From`. How you do that is up to you (and could involve something outside R), but you can use `select()` with or without the minus sign to include or exclude certain columns. You also can use the `rename` function to, well, rename columns.

```{r}
party_registration <- read_csv("data/new_party_reg.csv") 
party_registration

party_registration <- party_registration |>
  rename(Voting = ...1)
```

```{r}
# Removals
party_removals <- read_csv("data/removals_by_parties.csv") 
party_removals 

party_removals <- party_removals |>
  filter(!is.na(DEM)) |> 
  rename(Methods = ...1)
```

```{r}
# Changes
county_changes <- read_csv("data/county_changes.csv") 
county_changes

county_changes <- county_changes |>
  rename(Counties = ...1)
```

```{r}
# Current registrations
Current_regisrations <- read_csv("data/total_register.csv") 
Current_regisrations

Current_regisrations <- Current_regisrations |>
  rename(Counties = ...1) |>
select(-`DEM...4`, - `REP...5`, - `NLM...6`, -`LIB...7`, - `GRN...8`, - `UNAF...9`, - `OTH...10`, - `TOTAL...11`)
```

## Rinse, Repeat: September 2022 data

Do all of that again, for the [September 2022 voter registration update](https://elections.maryland.gov/pdf/vrar/2022_09.pdf), so we can compare the two. If you've done well, this is easy: just do the same steps in Tabula, export the files and use the same code to read in and clean up the data, *just remember to give these dataframes different names* so that you can compare September 2024 to September 2022.

To answer the following questions, you'll need to calculate percentages and do other calculations. Whether you do that by combining or joining dataframes or you write code to run separately on both is up to you.

```{r}
# Party registrations
party2022_registration <- read_csv("data/new_reg_2022.csv") 
party2022_registration

party2022_registration <- party2022_registration |>
  rename(Voting = ...1)
```

```{r}
# Removals
party2022_removals <- read_csv("data/removals_2022.csv") 
party2022_removals 

party2022_removals <- party2022_removals |>
  filter(!is.na(DEM)) |> 
  rename(Methods = ...1)
```

```{r}
# Changes
county2022_changes <- read_csv("data/county_changes_2022.csv") 
county2022_changes

county2022_changes <- county2022_changes |>
  rename(Counties = ...1)
```

```{r}
# Current registrations
the2022_regisrations <- read_csv("data/total_reg_2022.csv") 
the2022_regisrations

the2022_regisrations <- the2022_regisrations |>
  rename(Counties = ...1) |>
select(-`DEM...4`, - `REP...5`, - `GRN...6`, -`LIB...7`, - `WCP...8`, - `UNAF...9`, - `OTH...10`, - `TOTAL...11`)
```

## Answer questions

Q1. Calculate the percentage of registered Democrats, Republicans and unaffiliated voters for the entire state in September 2024 and September 2022, and then calculate the percentage change for those three party categories (the formula for percentage change is easy to remember: (New-Old)/Old). How would you describe the changes in these partisan categories in a story? What's the most interesting thing about these changes that would be the lede of that story?

A1. Democratic registration decreased the most between 2022 and 2024 between the two major parties, at over a three percent drop. Republican registration dropped off marginally, falling by .63% between 2022 and 2024. The most interesting change was unaffiliated voters registered at a much higher rate in 2024 than 2022. There was an eight percent increase of regisration of voters who were not affilated with either the Republican or Democratic party.

```{r}
the2022_percentages <- the2022_regisrations |>
  summarise(
    Dem_Percentage_2022 = sum(`DEM...12`) / sum(`TOTAL...19`) * 100,
    Rep_Percentage_2022 = sum(`REP...13`) / sum(`TOTAL...19`) * 100,
    Unaffiliated_Percentage_2022 = (sum(`TOTAL...19`) - sum(`DEM...12`) - sum(`REP...13`)) / sum(`TOTAL...19`) * 100
  )

Current_percentages <- Current_regisrations %>%
  summarise(
    Dem_Percentage_2024 = sum(`DEM...12`) / sum(`TOTAL...19`) * 100,
    Rep_Percentage_2024 = sum(`REP...13`) / sum(`TOTAL...19`) * 100,
    Unaffiliated_Percentage_2024 = (sum(`TOTAL...19`) - sum(`DEM...12`) - sum(`REP...13`)) / sum(`TOTAL...19`) * 100
  )

percentage_change <- tibble(
  Dem_Percentage_Change = (Current_percentages$Dem_Percentage_2024 - the2022_percentages$Dem_Percentage_2022) / the2022_percentages$Dem_Percentage_2022 * 100,
  Rep_Percentage_Change = (Current_percentages$Rep_Percentage_2024 - the2022_percentages$Rep_Percentage_2022) / the2022_percentages$Rep_Percentage_2022 * 100,
  Unaffiliated_Percentage_Change = (Current_percentages$Unaffiliated_Percentage_2024 - the2022_percentages$Unaffiliated_Percentage_2022) / the2022_percentages$Unaffiliated_Percentage_2022 * 100
)

```

Q2. Among the Maryland jurisdictions with the most registered voters, which one had the biggest percentage increase in unaffiliated voters between 2022 and 2024? How did that county vote for governor in 2022 and for president in 2020? Contrast the partisan percentages of that county with the county that had the eighth-largest percentage increase in unaffiliated voters. You should display information from the data that shows the counties with the largest percentage increase first.

A2. Prince George's county saw the highest increase in unaffiliated voters between 2022 and 2024, rising over 16 percent. PG voted for Wes Moore for governor in 2022 by a wide margin, granting him 89% of the counties vote. The county voted for President Joe Biden in 2020 as well, granting him just over 89% of the votes. Wicomico, the county with the eight-largest percentage increase in unaffiliated voters, voted for Donald Trump at 49 percent, compared to President Biden at 47%. 

```{r}
county2022_data <- the2022_regisrations |>
  mutate(Total_Unaffiliated_2022 = TOTAL...19 - (DEM...12 + REP...13))

county2024_data <- Current_regisrations |>
  mutate(Total_Unaffiliated_2024 = TOTAL...19 - (DEM...12 + REP...13))

county_data <- county2022_data |>
  select(Counties, Total_Unaffiliated_2022) |>
  inner_join(county2024_data |>
               select(Counties, Total_Unaffiliated_2024), 
             by = "Counties")

county_data <- county_data |>
  mutate(Unaffiliated_Increase = (Total_Unaffiliated_2024 - Total_Unaffiliated_2022) / Total_Unaffiliated_2022 * 100)
```

Q3. Do the party percentages of new registrants (% Dem, % Rep, % Unaffiliated) from September 2024 and September 2022 match the overall percentage of partisan registration in the state at those times?

In other words, did new registrants mirror the overall partisan composition of the electorate, or were they different? Do any of the methods of registration vary significantly from the overall partisan pattern? Which method is worthy of further exploration?

A3. In 2022, unaffiliated voters are registering at a much higher rate than the total party percentages. The percentage of new unaffiliated registrants is over 30%, while the total state is just at 22%. I think this shows that there is going to be more and more unaffiliated voters who register. That was similar to the registration in 2024, as the number of unaffiliated voters who registered was much higher than the state's total percentage. Motor vehicle registration also increased a large amount between 2022 and 2024, which offers some intrigue.

```{r}
new_2022_reg <- party2022_registration |>
  summarise(
    Dem_Percentage_2022 = sum(`DEM`) / sum(`TOTAL`) * 100,
    Rep_Percentage_2022 = sum(`REP`) / sum(`TOTAL`) * 100,
    Unaffiliated_Percentage_2022 = (sum(`TOTAL`) - sum(`DEM`) - sum(`REP`)) / sum(`TOTAL`) * 100
  )

new_2024_reg <- party_registration %>%
  summarise(
    Dem_Percentage_2024 = sum(`DEM`) / sum(`TOTAL`) * 100,
    Rep_Percentage_2024 = sum(`REP`) / sum(`TOTAL`) * 100,
    Unaffiliated_Percentage_2024 = (sum(`TOTAL`) - sum(`DEM`) - sum(`REP`)) / sum(`TOTAL`) * 100
  )
```

Q4. Choose your own adventure! Using both the 2024 and 2022 data for one of the dataframes (changes, removals or new registrations), ask and answer a question that makes at least one comparison and requires code to answer. That could involve calculating a percentage, percentage change or the raw difference between two columns.

A4. There is more people leaving there counties and the state of Maryland in 2024 then in 2022. That's interesting considering the rise in unaffiliated voters.

```{r}
removals_2024 <- party_removals |>
filter(Methods %in% c("COUNTY TRANSFER - OUT", "MOVED OUT OF STATE")) 

removals_2022 <- party2022_removals |>
filter(Methods %in% c("COUNTY TRANSFER - OUT", "MOVED OUT OF STATE")) 

removals_bothyears <- removals_2022 |>
  mutate(
    Percent_Removals_2022 = (16558 / 4161637) * 100,
    Percent_Removals_2024 = (12658 / 4231403) * 100,
  )

```

Q5. What's the best story idea or question you've seen as a result of the work you've done in this lab?

A5. Why are so many unaffiliated voters registering at a higher rate than the state's total population? Are people frustrated with the two-party system? Or is this that younger people are more likely to vote for neither party?
