---
title: "lab_04"
author: "Sam Jane"
date: "2024-09-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(janitor)
library(lubridate)

```

For this lab, we want to investigate spending by the two leading Senate candidates in Maryland during this election, Angela Alsobrooks and Larry Hogan. Our goal is to try and standardize the purpose of the expenditures so that we can compare the two campaigns. We have the data, but it's messy - purpose descriptions are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis.

## Load Data

You'll need to load one data set, a CSV file of campaign expenditures located in the data folder called "md_senate_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table
maryland_expenses <- read_csv("data/md_senate_expenses.csv")

#Display it
maryland_expenses
```

## Answer questions

**Q1.** You've been assigned to compare the spending priorities of the two campaigns, but they describe their spending differently using the `purpose` column. Using OpenRefine, create a project using the original CSV file, make a copy of the `purpose` column called `purpose_clean` and then standardize it, focusing on the purposes that appear most often or have the largest total amounts. You can transform the data as you want - making all records the same case, for example, to help. The goal is to be able to group together as many similar purpose values, so you can choose to call all transportation-related expenses the same thing.

How do you know when you're done? It's a judgment call, but you should be close to 100 unique values for `purpose_clean`, and maybe even below that number.

Then, when you are done standardizing `purpose_clean`, export it as a new CSV file to your data folder, *giving it a different name* and read it back into RStudio as a new dataframe.

Then, using that `purpose_clean` column, write code to find the largest amounts of spending on categories by both campaigns.

**A1.** 
#I found it to be interesting that  Hogan and Alsobrooks determine different spendings to be under different categories. But I found it fascinating that Alsobrooks was spending much more money on advertising (3,639,344.50) while Hogan was spending less (2,145,145.48). This is probably due to Hogan feeling like he's much more recognizable than ALsobrooks, because of his prior history.

```{r}
# Load the Maryland expenditures data table
maryland_expenses_clean <- read_csv("data/md-senate-expenses-clean.csv")

#Display it
maryland_expenses_clean
```
```{r}
Hogan_expenses <- maryland_expenses_clean |> 
filter (candidate=="Hogan") |> 
group_by(purpose_clean) |>
summarize (total = sum(amount)) |> 
arrange(desc(total))
```  

```{r}
Alsobrooks_expenses <- maryland_expenses_clean |> 
filter (candidate=="Alsobrooks") |> 
group_by(purpose_clean) |>
summarize (total = sum(amount)) |> 
arrange(desc(total))
```  
```



```

**Q2.** Let's do some research on the top purposes from A1. Feel free to search the Web to understand those terms better if need be, and then write a description comparing and contrasting the two campaigns' largest categories of spending.

**A2.**
# One of the elements I found interesting was the difference in spending between print and online for Alsobrooks and Hogan. Some of Hogan's top categories were in print, such as print ads, print services, etc. Meanwhile, one of Alsobrook's top spending was in online advertising. This is a very interesting detail, because I think it shows the changing coaltion of the voters that the candidates are targeting.

**Q3.** Pick one of the top categories for each campaign and write code that shows how that spending has changed over time, if at all. You'll need to use lubridate for this. Write a couple of sentences describing that pattern.

**A3.**
#Hogan only spent in the months of March, April, May and June, while Alsobrooks was spending the entire year of 2023.
```{r}
Hogan_fundraising <- maryland_expenses_clean |> 
  filter(candidate == "Hogan", purpose_clean == "FUNDRAISING SERVICES") |> 
  mutate(year = year(date), month = month(date)) |> 
  group_by(year, month) |> 
  summarize(total = sum(amount), .groups = "drop") |> 
  arrange(year, month)
```
```{r}
Alsobrooks_fundraising <- maryland_expenses_clean |> 
  filter(candidate == "Alsobrooks", purpose_clean == "Online Advertising") |> 
  mutate(year = year(date), month = month(date)) |> 
  group_by(year, month) |> 
  summarize(total = sum(amount), .groups = "drop") |> 
  arrange(year, month)
```

**Q4.** Which committee has the larger payroll or salary spending so far? How many unique recipients of payroll/staff spending are there for each candidate?

**A4.** 
#Alsobrooks has the higher amount of spending. There are a good majority of unique recipients.

```{r}
USE OTHER CODE
```

**Q5** What's the most interesting story idea you've found in this data?

**A5** 
#That ALsobrooks is spending much more on online advertising, while Hogan is focusing on TV and print it seems. 
