---
title: "lab_10"
author: "Sam Jane"
date: "2024-11-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   the tidyverse, janitor and tigris libraries
-   you might choose to use the tidycensus library for Q3

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load libraries
library(tidyverse)
library(sf)
# install.packages('tigris')
library(tigris)
library(janitor)
```

Let's explore the election results from last night and make some maps!

## Questions

**Q1.** Make a county-level map of Maryland's results for U.S. Senate using `md_senate_county_24.csv` in the data folder, calculating the difference between Angela Alsobrooks's percentage of the total votes and Larry Hogan's percentage of the total votes and displaying that percentage difference on the map. What is the story here, and what is the county with the most interesting results?

**A1.**
#The story is that there was a lot of close(ish) races in the senate Maryland election, but Alsobrooks signficantly overperformed, and was not really blown out in any county. The county with the most interesting results is Prince George's, which overwhelmingly voted for Alsobrooks, which makes sense considering her background there and also that she's a black woman, but the fact that she did so well in one of Maryland's largest populations despite some of the struggles she's had running the county there is interesting.

```{r}
md_senate_data <- read_csv("data/md_senate_county_24.csv") |>
  clean_names()
md_senate_data <- md_senate_data |>
  mutate(
    GEOID = as.character(geoid),
    total_votes = alsobrooks + hogan,  
    alsobrooks_percentage = (alsobrooks / total_votes) * 100,
    hogan_percentage = (hogan / total_votes) * 100,
    percent_diff = alsobrooks_percentage - hogan_percentage )
counties_md <- counties(state = "MD", class = "sf")
md_senate_map_data <- counties_md |>
  left_join(md_senate_data, by = c("GEOID"))
ggplot(data = md_senate_map_data) +
  geom_sf(aes(fill = percent_diff), color = "white") +
  scale_fill_gradient2(
    low = "purple", mid = "blue", high = "white",
    midpoint = 0,
  ) +
  labs(
    title = "Percentage Difference in Votes for U.S. Senate (Alsobrooks vs. Hogan)",
    subtitle = "Maryland Counties",
    caption = "Data source: md_senate_county_24.csv"
  ) +
  theme_minimal()
```

**Q2.** Make a county-level map showing the difference between Donald Trump's county-level performance this year and Larry Hogan's, using percentages to compare the two as you did in Q1. You can use the dataframe you initially loaded for Q1, and you'll need to load the 2024 presidential results in the data folder and join them to it before proceeding. Are there any counties where Trump got a higher percentage than Hogan? How would you describe the map showing the Trump-Hogan difference?

Also answer this: is a map the best way to present this data? What else could you make that might convey more information?

**A2.**
#There were no counties where Trump did better than Hogan, which makes sense with Maryland's blue state. I think it shows though that Trump did better than Maryland than I expected, because the percent difference was very low between the two candidates and I thought that there would be alot more split ticket voters to Hogan and not to Trump. That did not seem to be the case. I think a map is the best way to convey this, but a line graph in total votes might be interesting to compare 2020 to 2024.

```{r}
hogan_data <- read_csv("data/md_senate_county_24.csv") |> 
  clean_names()
trump_data <- read_csv("data/md_pres_county_24.csv") |> 
  clean_names()

hogan_data <- hogan_data |> 
  mutate(
    GEOID = as.character(geoid),
    total_votes_hogan = alsobrooks + hogan,
    hogan_percentage = (hogan / total_votes_hogan) * 100
  )

trump_data <- trump_data |> 
  mutate(
    GEOID = as.character(geoid),  
    total_votes_trump = harris + trump,  
    trump_percentage = (trump / total_votes_trump) * 100
  )

percent_data <- hogan_data |> 
  select(GEOID, hogan_percentage) |> 
  left_join(
    trump_data |> select(GEOID, trump_percentage), 
    by = "GEOID"
  ) |> 
  mutate(
    percent_diff = trump_percentage - hogan_percentage
  )

counties_md <- counties(state = "MD", class = "sf")

map_data <- counties_md |> 
  left_join(percent_data, by = c("GEOID"))

ggplot(data = map_data) +
  geom_sf(aes(fill = percent_diff), color = "white") +
  scale_fill_gradient2(
    low = "red", mid = "gray", high = "purple",
    midpoint = 0,
    name = "Trump - Hogan (%)"
  ) +
  labs(
    title = "Larry Hogan signficantly outperformed Donald Trump in Maryland",
    caption = "Data sources: md_senate_county_24.csv and presidential_results_2024.csv"
  ) +
  theme_minimal()
```

**Q3** Make another map showing the difference between Larry Hogan's county-level performance this year and from his governor race in 2018, using percentages to compare the two as you did in Q2. You can use the dataframe you initially loaded for Q1, and you'll need to load the 2018 governor results in the data folder and join them to it before proceeding. Are there any counties where Hogan did better this year? How would you describe the map showing the difference?

**A3**

```{r}
hogan_2018_data <- read_csv("data/md_gov_county_18.csv") |>
  clean_names()

hogan_2018_data <- hogan_2018_data |> 
  mutate(
    GEOID = as.character(geoid),
     total_votes_2018 = jealous + hogan + quinn + schlakman,
    hogan_2018_percentage = (hogan / total_votes_2018) * 100
  )

comparison_data <- hogan_data |> 
  select(GEOID, hogan_percentage) |> 
  left_join(
    hogan_2018_data |> select(GEOID, hogan_2018_percentage), 
    by = "GEOID"
  ) |> 
  mutate(
    percent_diff = hogan_percentage - hogan_2018_percentage
  )
counties_md <- counties(state = "MD", class = "sf")
map_data <- counties_md |> 
  left_join(comparison_data, by = c("GEOID"))

# Step 7: Create the map
ggplot(data = map_data) +
  geom_sf(aes(fill = percent_diff), color = "white") +
  scale_fill_gradient2(
    low = "red", mid = "black", high = "blue",
    midpoint = 0,
    name = "2024 - 2018 (%)"
  ) +
  labs(
    title = "Change in Larry Hogan's Vote Percentage: 2024 vs. 2018 Governor Race",
    subtitle = "County-Level Performance in Maryland",
    caption = "Data sources: hogan_data (2024) and governor_results_2018.csv"
  ) +
  theme_minimal()
```


**Q4.** Choose your own map adventure! In the data folder I've included Maryland county-level results for the abortion rights amendment vote this year, plus the 2020 presidential results by county and 2022 governor's results by county. Using one of those and at least one other contest for comparison, make a county-level map showing what you think is the most interesting aspect of that data, and describe the results in terms of a possible story. You can bring in other data, such as Census information, if you like.

**A4.**
# I think this shows that while there defintely was people who voted for abortion rights and Hogan, that difference was much less than I thought. This was one of the questions I had, and it showed that abortion was clearly a difference maker for voters.

```{r}
abortion_data <- read_csv("data/md_abortion_24.csv") |> 
  clean_names()

abortion_data <- abortion_data |> 
  rename(
    yes = `for`, 
    no = against
  )

abortion_data <- abortion_data |> 
  mutate(
    total_votes = yes + no,
    abortion_percentage = ( yes / total_votes) * 100
  )

percent_data_abortion <- hogan_data |> 
  select(county, hogan_percentage) |> 
  left_join(
    abortion_data |> select(county, abortion_percentage), 
    by = "county"
  ) |> 
  mutate(
    percent_diff_abortion_hogan = abortion_percentage - hogan_percentage
  )

map_data_abortion <- counties_md |> 
  left_join(percent_data_abortion, by = c("NAME" = "county"))

ggplot(data = map_data_abortion) +
  geom_sf(aes(fill = percent_diff_abortion_hogan), color = "white") +
  scale_fill_gradient2(
    low = "red", mid = "gray", high = "purple",
    midpoint = 0,
    name = "Abortion Rights - Hogan (%)"
  ) +
  labs(
    title = "Difference in Voters for Abortion Rights and Larry Hogan in Maryland",
    caption = "Data sources: md_senate_county_24.csv and md_abortion_county_24.csv"
  ) +
  theme_minimal()
```

-30-
